# Hypothesis Testing with Known Distributions {#HYPTESTDIST}


## Objectives

1) Know and properly use the terminology of a hypothesis test: *permutation test*, *exact test*, *null hypothesis*, *alternative hypothesis*, *test statistic*, *p-value*, and *power*.   

2) Conduct all four steps of a hypothesis test using probability models.  


## Hypothesis testing using probability models  

As a lead into the Central Limit Theorem in \@ref(HYPTESTCLT) and mathematical sampling distributions, we will look at a class of hypothesis testing where the null hypothesis specifies a probability model. In some cases, we can get an exact answer, and in others, we will use simulation to get an empirical p-value. By the way, a **permutation test** is an **exact test**; by this we mean we are finding all possible permutations in the calculation of the p-value. However, since the complete enumeration of all permutations is often difficult, we approximate it with randomization, simulation. Thus, the p-value from a randomization test is an approximation of the exact (permutation) test.  

Let's use three examples to illustrate the ideas of this chapter.

## Tappers and listeners

Here's a game you can try with your friends or family. Pick a simple, well-known song. Tap that tune on your desk, and see if the other person can guess the song. In this simple game, you are the tapper, and the other person is the listener.

A Stanford University graduate student named Elizabeth Newton conducted an experiment using the tapper-listener game.^[This case study is described in http://www.openintro.org/redirect.php?go=made-to-stick&redirect=simulation_textbook_pdf_preliminary Made to Stick by Chip and Dan Heath.]  In her study, she recruited 120 tappers and 120 listeners into the study. About 50\% of the tappers expected that the listener would be able to guess the song. Newton wondered, is 50\% a reasonable expectation?

### Step 1- State the null and alternative hypotheses  

Newton's research question can be framed into two hypotheses:  

    $H_0$: The tappers are correct, and, in general, 50\% of listeners are able to guess the tune. $p = 0.50$    

    $H_A$: The tappers are incorrect, and either more than or less than 50\% of listeners are able to guess the tune. $p \neq 0.50$  


>**Exercise**:
Is this a one-sided or two-sided hypothesis test? How many variables are in this model?

The tappers think that listeners will guess the song correctly 50\% of the time, and this is a two-sided test since we don't know beforehand if listeners will be better or worse than 50\%.  

There is only one variable of interest, whether the listener is correct.


### Step 2 - Compute a test statistic 

In Newton's study, only 42 (we changed the number to make this problem more interesting from an educational perspective) out of 120 listeners ($\hat{p} = 0.35$) were able to guess the tune! From the perspective of the null hypothesis, we might wonder, how likely is it that we would get this result from chance alone? That is, what's the chance we would happen to see such a small fraction if $H_0$ were true and the true correct-guess rate is 0.50?

Now before we use simulation, let's frame this as a probability model. The random variable $X$ is the number of correct guesses out of 120. If the observations are independent and the probability of success is constant (each listener has the same probability of guessing correctly), then we could use a binomial model. We can't assess the validity of these assumptions without knowing more about the experiment, the subjects, and the data collection. For educational purposes, we will assume they are valid. Thus, our test statistic is the number of successes in 120 trials. The observed value is 42.

### Step 3 - Determine the p-value

We now want to find the p-value as $2 \cdot \Prob(X \leq 42)$ where $X$ is a binomial random variable with $p = 0.5$ and $n = 120$. Again, the p-value is the probability of the observed data or something more extreme, given the null hypothesis is true. Here, the null hypothesis being true implies that the probability of success is 0.50. We will use `R` to get the one-sided p-value and then double it to get the two-sided p-value for the problem. We selected $\Prob(X \leq 42)$ because "more extreme" means the observed values and values further from what you would get if the null hypothesis were true, which is 60 for this problem.

```{r}
2*pbinom(42, 120, prob = 0.5)
```

That is a small p-value.

### Step 4 - Draw a conclusion 

Based on our data, if the listeners were guessing correct 50\% of the time, there is less than a $0.0013$ probability that only 42 or less (or 78 or more) listeners would get correctly. This is the probability of what we observed or something more extreme, given the null hypothesis is true. This probability is much less than 0.05, so we reject the null hypothesis that the listeners are guessing correctly half of the time and conclude that the correct-guess rate rate is different from 50\%. 

This decision region looks like the pmf in Figure \@ref(fig:dist201-fig). Any observed values inside the red boundary lines would be consistent with the null hypothesis. That is, any observed values inside the red boundary lines would result in a p-value larger than 0.05. Any values at the red line or more extreme would be in the rejection region, resulting in a p-value smaller than 0.05. We also plotted the observed value in black.

```{r dist201-fig,warning=FALSE,fig.cap="Binomial pmf"}
gf_dist("binom", size = 120, prob = 0.5, xlim = c(50, 115)) %>%
  gf_vline(xintercept = c(48, 72), color = "red") %>%
  gf_vline(xintercept = c(42), color = "black") %>%
  gf_theme(theme_bw) %>%
  gf_labs(title = "Binomial pmf", subtitle = "Probability of success is 0.5", 
          y = "Probability")
```


### Repeat using simulation  

We will repeat the analysis using an empirical (observed from simulated data) p-value. Step 1, stating the null and alternative hypothesis, is the same.

### Step 2 - Compute a test statistic 

We will use the proportion of listeners that get the song correct instead of the number. This is a minor change since we are simply dividing by 120. 

```{r}
obs <- 42 / 120
obs
```


### Step 3 - Determine the p-value

To simulate 120 games under the null hypothesis where $p = 0.50$, we could flip a coin 120 times. Each time the coin comes up heads, this could represent the listener guessing correctly, and tails would represent the listener guessing incorrectly. For example, we can simulate 5 tapper-listener pairs by flipping a coin 5 times:  

$$
\begin{array}{ccccc} 
H & H & T & H & T \\
Correct & Correct & Wrong & Correct & Wrong \\
\end{array} 
$$

After flipping the coin 120 times, we got 56 heads for a proportion of $\hat{p}_{sim} = 0.467$. As we did with the randomization technique, seeing what would happen with one simulation isn't enough. In order to evaluate whether our originally observed proportion of 0.35 is unusual or not, we should generate more simulations. Here, we've repeated this simulation 10,000 times:

```{r}
set.seed(604)
results <- rbinom(10000, 120, 0.5) / 120
```

Note, we could simulate it a number of ways. Here is a way using `do()` that will look like what we've done for other randomization tests.

```{r cache=TRUE}
set.seed(604)
results <- do(10000)*mean(sample(c(0, 1), size = 120, replace = TRUE))
```

```{r}
head(results)
```


```{r dens202-fig,warning=FALSE,fig.cap="The estimated sampling distribution."}
results %>%
  gf_histogram(~mean, fill = "cyan", color = "black") %>%
  gf_vline(xintercept = c(obs, 1 - obs), color = "black") %>%
  gf_theme(theme_bw()) %>%
  gf_labs(x = "Test statistic")
```

Notice in Figure \@ref(fig:dens202-fig) how the sampling distribution is centered at 0.5 and looks symmetrical.

The p-value is found using the `prop1` function. In this problem, we really need the observed value to be included to prevent a p-value of zero.

```{r}
2*prop1(~(mean <= obs), data = results)
```

### Step 4 - Draw a conclusion 

In these 10,000 simulations, we see very few results close to 0.35. Based on our data, if the listeners were guessing correct 50\% of the time, there is less than a $0.0012$ probability that only 35% or less or 65% or more listeners would get it right. This p-value is much less than 0.05, so we reject that the listeners are guessing correctly half of the time and conclude that the correct-guess rate is different from 50\%. 

>**Exercise**:
In the context of the experiment, what is the p-value for the hypothesis test?^[The p-value is the chance of seeing the observed data or something more in favor of the alternative hypothesis (something as or more extreme) given that guessing has a probability of success of 0.5. Since we didn't observe many simulations with even close to just 42 listeners correct, the p-value will be small, around 1-in-1000.]

>**Exercise**:  
Do the data provide statistically significant evidence against the null hypothesis? State an appropriate conclusion in the context of the research question.^[The p-value is less than 0.05, so we reject the null hypothesis. There is statistically significant evidence, and the data provide strong evidence that the chance a listener will guess the correct tune is different from 50\%.]

## Cardiopulmonary resuscitation (CPR)

Let's return to the CPR example from last chapter. As a reminder, we will repeat some of the background material.

Cardiopulmonary resuscitation (CPR) is a procedure sometimes used on individuals suffering a heart attack. It is helpful in providing some blood circulation to keep a person alive, but CPR chest compressions can also cause internal injuries, which complicate additional treatment efforts. For instance, blood thinners may be used to help release a clot that is causing the heart attack, but blood thinners negatively affect internal injuries.

Patients who underwent CPR for a heart attack and were subsequently admitted to a hospital^["Efficacy and safety of thrombolytic therapy after initially unsuccessful cardiopulmonary resuscitation: a prospective clinical trial." The Lancet, 2001.] were randomly assigned to either receive a blood thinner (treatment group) or not receive a blood thinner (control group). The outcome variable of interest was whether the patient survived for at least 24 hours.

### Step 1- State the null and alternative hypotheses  

We want to understand whether blood thinners are helpful or harmful. We'll consider both of these possibilities using a two-sided hypothesis test.

$H_0$: Blood thinners do not have an overall survival effect; survival rate is independent of experimental treatment group. $p_c - p_t = 0$.  
$H_A$: Blood thinners have an impact on survival, either positive or negative, but not zero. $p_c - p_t \neq 0$.  

```{r message=FALSE,warning=FALSE}
thinner <- read_csv("data/blood_thinner.csv")
```

```{r}
head(thinner)
```


Let's put the data in a table.

```{r}
tally(~group + outcome, data = thinner, margins = TRUE)
```

### Step 2 - Compute a test statistic. 

In this example, we can think of the data as coming from a **hypergeometric** distribution. This is really a binomial from a finite population. We can calculate the p-value using this probability distribution. The random variable is the number of control patients that survived from a total population of 90 patients, where 50 are control patients and 40 are treatment patients, and where a total of 25 survived.

### Step 3 - Determine the p-value.

In this case, we want to find $\Prob(X \leq 11)$ (the observed number of control patients that survived) and double it since it is a two-sided test.

```{r}
2*phyper(11, 50, 40, 25)
```

Note: We could have picked the lower right cell as the reference cell. But now I want the $\Prob(X \geq 14)$ (the observed number of treatment patients that survived) with the appropriate change in parameter values. Notice we get the same answer.

```{r}
2*(1 - phyper(13, 40, 50, 25))
```

We could do the same thing for the other two cells. Here we find $\Prob(X \leq 26)$ (the observed number of treatment patients that did not survive).

```{r}
2*phyper(26, 40, 50, 65)
```

Here we find $\Prob(X \geq 39)$ (the observed number of control patients that did not survive). 

```{r}
2*(1 - phyper(38, 50, 40, 65))
```


`R` also has a built in function, `fisher.test()`, that we could use. This function calculates Fisher's exact test, where p-values are obtained using the hypergeometric distribution. 

```{r}
fisher.test(tally(~group + outcome, data = thinner))
```

The p-value is slightly different since the **hypergeometric** distribution is not symmetric. For this reason, doubling the p-value from the single side result is not quite right. The algorithm in `fisher.test()` finds and adds all probabilities less than or equal to value of $\Prob(X = 11)$, see Figure \@ref(fig:dens203-fig). Using `fisher.test()` gives the correct p-value.

```{r dens203-fig,fig.cap="Hypergeometric pmf showing the cutoff for p-value calculation.",warning=FALSE}
gf_dist("hyper", m = 50, n = 40, k = 25) %>%
  gf_hline(yintercept = dhyper(11, 50, 40, 25), color = "red") %>%
  gf_labs(title = "Hypergeometric pmf", subtitle = "Red line is P(X = 11)", 
          y = "Probability") %>%
  gf_theme(theme_bw())
```

This is how `fisher.test()` is calculating the p-value:

```{r}
temp <- dhyper(0:25, 50, 40, 25)
sum(temp[temp <= dhyper(11, 50, 40, 25)])
```


The randomization test in the last chapter yielded a p-value of 0.257 so all tests are consistent.

### Step 4 - Draw a conclusion 

Since this p-value is larger than 0.05, we fail to reject the null hypothesis. That is, we do not find statistically significant evidence that the blood thinner has any influence on survival of patients who undergo CPR prior to arriving at the hospital. Once again, we can discuss the causal conclusion since this is an experiment.


Notice that in these first two examples, we had a test of a single proportion and a test of two proportions. The single proportion test did not have an equivalent randomization test since there is not a second variable to shuffle. We were able to get answers since we found a probability model that we could use instead.  

## Golf Balls  

Our last example will be interesting because the distribution has multiple parameters and a test metric is not obvious at this point.

The owners of a residence located along a golf course collected the first 500 golf balls that landed on their property. Most golf balls are labeled with the make of the golf ball and a number, for example "Nike 1" or "Titleist 3". The numbers are typically between 1 and 4, and the owners of the residence wondered if these numbers are equally likely (at least among golf balls used by golfers of poor enough quality that they lose them in the yards of the residences along the fairway.)

We will use a significance level of $\alpha = 0.05$ since there is no reason to favor one decision error over the other.

### Step 1- State the null and alternative hypotheses  

We think that the numbers are not all equally likely. The question of one-sided versus two-sided is not relevant in this test. You will see this when we write the hypotheses. 

$H_0$: All of the numbers are equally likely.$\pi_1 = \pi_2 = \pi_3 = \pi_4$ Or $\pi_1 = \frac{1}{4}, \pi_2 =\frac{1}{4}, \pi_3 =\frac{1}{4}, \pi_4 =\frac{1}{4}$  
$H_A$: There is some other distribution of the numbers in the population. At least one population proportion is not $\frac{1}{4}$.  

Notice that we switched to using $\pi$ instead of $p$ for the population parameter. There is no reason other than to make you aware that both are used. 

This problem is an extension of the binomial. Instead of two outcomes, there are four outcomes. This is called a multinomial distribution. You can read more about it if you like, but our methods will not make it necessary to learn the probability mass function.


Out of the 500 golf balls collected, 486 of them had a number between 1 and 4. We will deal with only these 486 golf balls. Let's get the data from `golf_balls.csv".

```{r echo=FALSE,eval=FALSE}
set.seed(1)
golf_balls <- tibble(number = rep(1:4, c(137, 138, 107, 104)))
golf_balls <- golf_balls %>% 
  sample() %>%
  select(-orig.id)
write_csv(golf_balls, "data/golf_balls.csv")
```


```{r message=FALSE,warning=FALSE}
golf_balls <- read_csv("data/golf_balls.csv")
```

```{r message=FALSE,warning=FALSE}
inspect(golf_balls)
```

```{r}
tally(~number, data = golf_balls)
```

### Step 2 - Compute a test statistic. 

If all numbers were equally likely, we would expect to see 121.5 golf balls of each number. This is a point estimate and thus, not an actual value that could be realized. Of course, in a sample we will have variation and thus, departure from this state. We need a test statistic that will help us determine if the observed values are reasonable under the null hypothesis. Remember that the test statistic is a single number metric used to evaluate the hypothesis.

>**Exercise**:  
What would you propose for the test statistic?  

With four proportions, we need a way to combine them. This seems tricky, so let's just use a simple approach. Let's take the maximum number of balls across all cells of the table and subtract the minimum. This is called the range and we will denote the parameter as $R$. Under the null hypothesis, this should be zero. We could re-write our hypotheses as: 

$H_0$: $R=0$  
$H_A$: $R>0$  

Notice that $R$ will always be non-negative, thus this test is one-sided.

The observed range is 34, $138 - 104$.

```{r}
obs <- diff(range(tally(~number, data = golf_balls)))
obs
```

### Step 3 - Determine the p-value.

We don't know the distribution of our test statistic, so we will use simulation. We will simulate data from a multinomial distribution under the null hypothesis and calculate a new value of the test statistic. We will repeat this 10,000 times and this will give us an estimate of the sampling distribution.

We will use the `sample()` function again to simulate the distribution of numbers under the null hypothesis. To help us understand the process and build the code, we are only initially using a sample size of 12 to keep the printout reasonable and easy to read.

```{r}
set.seed(3311)
diff(range(table(sample(1:4, size = 12, replace = TRUE))))
```

Notice this is not using `tidyverse` coding ideas. We don't think we need tibbles or data frames so we went with straight nested `R` code. You can break this code down by starting with the code in the center.

```{r}
set.seed(3311)
sample(1:4, size = 12, replace = TRUE)
```

```{r}
set.seed(3311)
table(sample(1:4, size = 12, replace = TRUE))
```

```{r}
set.seed(3311)
range(table(sample(1:4, size = 12, replace = TRUE)))
```


```{r}
set.seed(3311)
diff(range(table(sample(1:4, size = 12, replace = TRUE))))
```

We are now ready to ramp up to the full problem. Let's simulate the data under the null hypothesis. We are sampling 486 golf balls (instead of 12) with the numbers 1 through 4 on them. Each number is equally likely. We then find the range, our test statistic. Finally we repeat this 10,000 to get an estimate of the sampling distribution of our test statistic. 

```{r cache=TRUE}
set.seed(3311)
results <- do(10000)*diff(range(table(sample(1:4, size = 486, replace = TRUE))))
```

Figure \@ref(fig:dens204-fig) is a plot of the sampling distribution of the range.

```{r dens204-fig,warning=FALSE,fig.cap="Sampling distribution of the range."}
results %>%
  gf_histogram(~diff, fill = "cyan", color = "black") %>%
  gf_vline(xintercept = obs, color = "black") %>%
  gf_labs(title = "Sampling Distribution of Range", 
          subtitle = "Multinomial with equal probability",
          x = "Range") %>%
  gf_theme(theme_bw)
```

Notice how this distribution is skewed to the right. The p-value is 0.14. 

```{r}
prop1(~(diff >= obs), data = results)
```


### Step 4 - Draw a conclusion 

Since this p-value is larger than 0.05, we fail to reject the null hypothesis. That is, based on our data, we do not find statistically significant evidence against the claim that the number on the golf balls are equally likely. We can't say that the proportion of golf balls with each number differs from 0.25. 

## Repeat with a different test statistic

The test statistic we developed was helpful, but it seems weak because we did not use the information in all four cells. So let's devise a metric that does this. The hypotheses are the same, so we will jump to step 2.

### Step 2 - Compute a test statistic. 

If each number were equally likely, we would have 121.5 balls in each bin. We can find a test statistic by looking at the deviation in each cell from 121.5.

```{r}
tally(~number, data = golf_balls) - 121.5
```

Now we need to collapse these into a single number. Just adding will always result in a value of 0, why? So let's take the absolute value and then add the cells together.

```{r}
obs <- sum(abs(tally(~number, data = golf_balls) - 121.5))
obs
```

This will be our test statistic.  

### Step 3 - Determine the p-value.

We will use similar code from above with our new metric. Now we sample 486 golf balls with the numbers 1 through 4 on them, and find our test statistic, the sum of the absolute deviations of each cell of the table from the expected count, 121.5. We repeat this process 10,000 times to get an estimate of the sampling distribution of our test statistic. 

```{r cache=TRUE}
set.seed(9697)
results <- do(10000)*sum(abs(table(sample(1:4, size = 486, replace = TRUE)) - 121.5))
```

Figure \@ref(fig:dens205-fig) is a plot of the sampling distribution of the absolute value of deviations.

```{r dens205-fig,fig.cap="Sampling distribution of the absolute deviations.",warning=FALSE}
results %>%
  gf_histogram(~sum, fill = "cyan", color = "black") %>%
  gf_vline(xintercept = obs, color = "black") %>%
  gf_labs(title = "Sampling Distribution of Absolute Deviations",
          subtitle = "Multinomial with equal probability",
          x = "Absolute deviations") %>%
  gf_theme(theme_bw)
```

Notice how this distribution is skewed to the right and our test statistic seems to be more extreme.

```{r}
prop1(~(sum >= obs), data = results)
```

The p-value is 0.014. This value is much smaller than our previous result. The test statistic matters in our decision process as nothing about this problem has changed except the test statistic.


### Step 4 - Draw a conclusion 

Since this p-value is smaller than 0.05, we reject the null hypothesis. That is, based on our data, we find statistically significant evidence against the claim that the numbers on the golf balls are equally likely. We conclude that the numbers on the golf balls are not all equally likely, or that at least one is different. 

## Summary  

In this chapter, we used probability models to help us make decisions from data. This chapter is different from the randomization section in that randomization had two variables (one of which we could shuffle) and a null hypothesis of no difference. In the case of a single proportion, we were able to use the binomial distribution to get an exact p-value under the null hypothesis. In the case of a $2 \times 2$ table, we were able to show that we could use the hypergeometric distribution to get an exact p-value under the assumptions of the model.

We also found that the choice of test statistic has an impact on our decision. Even though we get valid p-values and the desired Type 1 error rate, if the information in the data is not used to its fullest, we will lose power. Note: **power** is the probability of correctly rejecting the null hypothesis when the alternative hypothesis is true.

In the next chapter, we will learn about mathematical solutions to finding the sampling distribution. The key difference in all these methods is the selection of the test statistic and the assumptions made to derive a sampling distribution.


## Homework Problems

1. Repeat the analysis of the yawning data from last chapter, but this time use the hypergeometric distribution.

    Is yawning contagious? 

    An experiment conducted by the *MythBusters*, a science entertainment TV program on the Discovery Channel, tested if a person can be subconsciously influenced into yawning if another person near them yawns. 50 people were randomly assigned to two groups: 34 to a group where a person near them yawned (treatment) and 16 to a group where there wasn't a person yawning near them (control). The following table shows the results of this experiment. 

$$
\begin{array}{cc|ccc} & & &\textbf{Group}\\ 
& & \text{Treatment } &  \text{Control} & \text{Total}  \\
& \hline \text{Yawn}	&	10	 	& 4		& 14  \\
\textbf{Result} & \text{Not Yawn}	& 24	 	& 12 	 	& 36   \\
	&\text{Total}		& 34		& 16		& 50 \\
\end{array} 
$$

The data is in the file `yawn.csv`.

a. What are the hypotheses?   

b. Choose a cell, and calculate the observed statistic.  

c. Find the p-value using the hypergeometric distribution.  

d. Plot the the sampling distribution.  

e. Determine the conclusion of the hypothesis test.  

f. Compare your results with the randomization test.  


2. Repeat the analysis of the golf ball data using a different test statistic.

    Use a level of significance of 0.05.

a. State the null and alternative hypotheses.  

b. Compute a test statistic.  

c. Determine the p-value.  

d. Draw a conclusion.  


3. Body Temperature

Shoemaker^[L. Shoemaker Allen (1996) What's Normal? – Temperature, Gender, and Heart Rate, Journal of Statistics Education, 4:2] cites a paper from the American Medical Association^[Mackowiak, P. A., Wasserman, S. S., and Levine, M. M. (1992), "A Critical Appraisal of 98.6 Degrees F, the Upper Limit of the Normal Body Temperature, and Other Legacies of Carl Reinhold August Wunderlich," Journal of the American Medical Association, 268, 1578-1580.] that questions conventional wisdom that the average body temperature of a human is 98.6 degrees Fahrenheit. One of the main points of the original article is that the traditional mean of 98.6 is, in essence, 100 years out of date. The authors cite problems with the original study's methodology, diurnal fluctuations (up to 0.9 degrees F per day), and unreliable thermometers. The authors believe the average temperature is less than 98.6. Conduct a hypothesis test.   

a. State the null and alternative hypotheses.  

b. State the significance level that will be used.  

c. Load the data from the file "temperature.csv" and generate summary statistics and a boxplot of the temperature data. We will not be using gender or heart rate for this problem.  

d. Compute a test statistic. We are going to help you with this part. We cannot do a randomization test since we don't have a second variable. It would be nice to use the mean as a test statistic but we don't yet know the sampling distribution of the sample mean.  

    Let's get clever. If the distribution of the sample is symmetric (this is an assumption but look at the boxplot and summary statistics to determine if you are comfortable with it), then under the null hypothesis, the observed values should be equally likely to either be greater or less than 98.6. Thus, our test statistic is the number of cases that have a positive difference between 98.6 and the observed value. This will be a binomial distribution with a probability of success (having a positive difference) of 0.5. You must also account for the possibility that there are observations of 98.6 in the data.

e. Determine the p-value.  

f. Draw a conclusion.  


